---
title: "DESCRIBING RELATIONSHIPS"
subtitle: "2. Cross&ndash;Tabulation"
author: "Austin Hart"
institute: "American University"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, rladies, rladies-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  message=FALSE, warning=FALSE, eval = TRUE, echo = FALSE, 
  fig.align = 'center', dev='svg'
)
```

```{r results='hide'}
library(tidyverse)
library(magrittr)
library(kableExtra)

film = readxl::read_excel("biopics.xls")
```


# Cross&ndash;Tabulation


Assessing the relationship between:  

- Nominal exposure variable (X)  

- Nominal outcome variable (Y)  


Data used: "biopics.xls"


---
# Basic tabulation revisited

.pull-left[

Women's representation among 761 biopic stars, 1915&ndash;2014.
  

```{r sumtab}
  tab = 
    film %>%
    count(SubjectSex) %>%
    mutate(
      Percent = 100 * n/sum(n)
    ) 

  tab %>%
    kbl(digits = 1, format = "html", caption = "Biopic Stars",
        col.names = c("","N","%")) %>%
    kable_paper(bootstrap_options = "striped", full_width = F, position = "center") %>%
    column_spec(3, bold = T)

```

]

.pull-right[

```{r boxsum, fig.width=3.5, fig.height=5, dpi=200}
  barplot(Percent ~ SubjectSex, data = tab,
        main = "Biopic Stars, 1915-2014", xlab = NULL,
        ylim = c(0,100), col="#69b3a2")
```

]


---
# Cross&ndash;Tabulation

> Does the distribution on the outcome (i.e. relative frequency table) differ by groups?

- Build a cross&ndash;table: groups define the columns!

- Compare the distribution by group


---
# Building a cross&ndash;table

### Representation by era

```{r sumtabgrp1}
  film %>%
    filter(Period == "1965--1999") %>%
    count(SubjectSex) %>%
    mutate(Percent = 100 * n/sum(n)) %>%
    kbl(digits = 1, format = "html", caption = "1965-1999",
        col.names = c("","n","%")) %>%
    kable_paper(bootstrap_options = "striped", full_width = F, position = "center") %>%
    column_spec(3, bold = T)

```


.pull-left[


```{r sumtabgrp3}
  film %>%
    filter(Period == "1915--1965") %>%
    count(SubjectSex) %>%
    mutate(Percent = 100 * n/sum(n)) %>%
    kbl(digits = 1, format = "html", caption = "1915-1964",
        col.names = c("","n","%")) %>%
    kable_paper(bootstrap_options = "striped", full_width = F, position = "center") %>%
    column_spec(3, bold = T)

``` 

]

.pull-right[

```{r sumtabgrp2}
  film %>%
    filter(Period == "2000--2014") %>%
    count(SubjectSex) %>%
    mutate(Percent = 100 * n/sum(n)) %>%
    kbl(digits = 1, format = "html", caption = "2000-2014",
        col.names = c("","n","%")) %>%
    kable_paper(bootstrap_options = "striped", full_width = F, position = "center") %>%
    column_spec(3, bold = T)

```

]


---
# A full cross&ndash;tabulation

### Change in representation over time?

```{r xtabs}
  xtab =
    film %>% 
    count(SubjectSex,Period) %>%  
    na.omit() %>%
    pivot_wider(
      names_from = Period,
      values_from = n, 
      values_fill = 0
    ) %>%
    mutate_if(
      is.integer, list(Percent = ~./sum(.) * 100)
    ) %>% 
    select(SubjectSex,sort(tidyselect::peek_vars()))

  xtab %>%
    kbl(digits = 1, format = "html", caption = "Biopic Stars by Era",
        col.names = c(" ","n","%","n","%","n","%")) %>%
    kable_paper(bootstrap_options = "striped", full_width = F, position = "center") %>%
    column_spec(c(3,5,7), bold = T) %>%
    add_header_above(c(" " = 1,"1915-64" = 2, "1965-99" = 2, "2000-14" = 2))
```


---
class: inverse

# Tabular analysis


- Assessing relationship between:  

  - Nominal xposure variable (X)  

  - Nominal outcome variable (Y)  

- Process: tabulate by group and compare!  
  
- More info  

  - R replication script  
  
  - R Survival Guide  
  
  - Textbook chapter
